---
title: "Chapter 7 Tutorial: Tidying Your Data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```


## Before You Start

### Objectives

Students will 


### Timeline

Day 1

  * Introduction and Objectives
  
  * Explore 3 Data Cleaning Topics
  
  * Summary, Resources, Quiz
  
Day 2

  * Quick Review and Q&A
  
  * Explore 3 Advanced Cleaning Topics
  
  * Summary, Resources, Quiz

### Getting Set Up

*All of the functions will be able to run in this tutorial without opening RStudio.  When you have finished the tutorial, you will want to try some of these techniques on your own data.  The functions we run for this tutorial are supported by the tidyr and dplyr packages.  Install and load these packages before tidying your own data.*


## Day One: Pivoting Your Data



## Day One: Summarizing Data



## Day One: Dealing with NA's



## Day One: Summary and Quiz

### Recap

### Resources

Data Wrangling Cheat Sheet

https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

### Follow-up Videos

![](https://www.youtube.com/watch?v=jOd65mR1zfw)

![](https://www.youtube.com/watch?v=1ELALQlO-yM)

### Quiz

*This quiz is not for a grade, but a helpful guide to what you know and things you might brush up on.*


```{r quiz}
quiz(
  question("Which format is your data in when the columns are variables and the rows are observations?",
    answer("short"),
    answer("wide"),
    answer("long", correct = TRUE),
    answer("tall")
  ),
  question("Which function helps to convert data from wide to long format  by when the user identifies a key and value?",
    answer("gather", correct = TRUE),
    answer("spread"),
    answer("transpose"),
    answer("reshape")
  ),
   question("Which command gives the basic descriptive stats and types for all the variables in a dataframe?",
    answer("summary", correct = TRUE),
    answer("summarize"),
    answer("stats"),
    answer("grid")
  ),
   question("Which command would denote including data that is not missing?",
    answer("!is.na", correct = TRUE),
    answer("!"),
    answer("is.na"),
    answer("na", correct = TRUE)
  ),
   question("Which of the R packages listed below are used to help tidy data?",
    answer("tidyr", correct = TRUE),
    answer("learnr"),
    answer("ggplot2"),
    answer("tidyverse", correct = TRUE)
  )
  
)
```

## Day Two: Merging Data

## Day Two: Dates

## Day Two: Times

## Day Two: Summary and Quiz

### Recap

### Resources

### Follow-up Videos

![](https://www.youtube.com/watch?v=Zc_ufg4uW4U)

### Quiz


## Session Information
```{r sessioninfo, echo=FALSE}

sessionInfo()

```

