---
title: "Chapter 7 Tutorial: Tidying Your Data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Before You Start

### Objectives


### Timeline

Day 1

  * Introduction and Objectives
  
  * Explore 3 Data Cleaning Topics
  
  * Summary, Resources, Quiz
  
Day 2

  * Quick Review and Q&A
  
  * Explore 3 Advanced Cleaning Topics
  
  * Summary, Resources, Quiz

### Getting Set Up

*All of the functions will be able to run in this tutorial without opening RStudio.  When you have finished the tutorial, you will want to try some of these techniques on your own data.  The functions we run for this tutorial are supported by the tidyr and dplyr packages.  Install and load these packages before tidying your own data.*


## Day One: Pivoting Your Data

## Day One: Summarizing Data

## Day One: Dealing with NA's

## Day One: Summary and Quiz

### Recap

### Resources

Data Wrangling Cheat Sheet

```{r cheatsheet, out.width="100%", out.height="500"}

knitr::include_graphics('www/data_wrangling_cheatsheet.pdf')

```

### Follow-up Videos

![](https://www.youtube.com/watch?v=jOd65mR1zfw)

![](https://www.youtube.com/watch?v=1ELALQlO-yM)

### Quiz

## Day Two: Merging Data

## Day Two: Dates

## Day Two: Times

## Day Two: Summary and Quiz

### Recap

### Resources

### Follow-up Videos

![](https://www.youtube.com/watch?v=Zc_ufg4uW4U)

### Quiz