---
title: "Chapter 7 Tutorial: Tidying Your Data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```


## Before You Start

### Objectives

Students will 


### Timeline

Day 1

  * Introduction and Objectives
  
  * Explore 3 Data Cleaning Topics
  
  * Summary, Resources, Quiz
  
Day 2

  * Quick Review and Q&A
  
  * Explore 3 Advanced Cleaning Topics
  
  * Summary, Resources, Quiz

### Getting Set Up

*All of the functions will be able to run in this tutorial without opening RStudio.  When you have finished the tutorial, you will want to try some of these techniques on your own data.  The functions we run for this tutorial are supported by the tidyr and dplyr packages.  Install and load these packages before tidying your own data.*


## Day One: Pivoting Your Data



## Day One: Summarizing Data



## Day One: Dealing with NA's



## Day One: Summary and Quiz

### Recap

### Resources

Data Wrangling Cheat Sheet

https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

### Follow-up Videos

![](https://www.youtube.com/watch?v=jOd65mR1zfw)

![](https://www.youtube.com/watch?v=1ELALQlO-yM)

### Quiz

*This quiz is not for a grade, but a helpful guide to what you know and things you might brush up on.*


```{r quiz}
quiz(
  question("Which format is your data in when the columns are variables and the rows are observations?",
    answer("short"),
    answer("wide"),
    answer("long", correct = TRUE),
    answer("tall")
  ),
  question("Which function helps to convert data from wide to long format  by when the user identifies a key and value?",
    answer("gather", correct = TRUE),
    answer("spread"),
    answer("transpose"),
    answer("reshape")
  ),
   question("Which command gives the basic descriptive stats and types for all the variables in a dataframe?",
    answer("summary", correct = TRUE),
    answer("summarize"),
    answer("stats"),
    answer("grid")
  ),
   question("Which command would denote including data that is not missing?",
    answer("!is.na", correct = TRUE),
    answer("!"),
    answer("is.na"),
    answer("na", correct = TRUE)
  ),
   question("Which of the R packages listed below are used to help tidy data?",
    answer("tidyr", correct = TRUE),
    answer("learnr"),
    answer("ggplot2"),
    answer("tidyverse", correct = TRUE)
  )
  
)
```

## Day Two: Merging Data

Imagine you are the regional manager of Winco in Idaho. One thing you might want to do is combine the data from Boise and Idaho Falls. Let's start with data on how many customers visited each location, on each day in 2020. 

### Examine each dataset
Start by looking at the Boise location and examine the dimensions.

```{r Win1, message=FALSE, warning=FALSE}
Winco1<-read.csv("./Messy_CSVs/MergeData/WincoMerge1.csv")[,-1]
```

```{r win1L, exercise=TRUE}
#Examine the Boise data set
head(Winco1)
```

Then look at the Idaho Falls location and examine the dimensions of that dataset too

```{r Win2, message=FALSE, warning=FALSE}
Winco2<-read.csv("./Messy_CSVs/MergeData/WincoMerge2.csv")[,-1]
```

```{r Win2L, exercise=TRUE}
#Examine the Idaho Falls data set
head(Winco2)
```

### Merge with `rbind`
We can see that these two datasets have the same number of columns and the same column names. This is the most simple way to merge data.

We can stack these two datasets on top of eachother using the `rbind` command.
```{r rbindreal, message=FALSE, warning=FALSE}
allWinDat<- rbind(Winco1,Winco2)
```

```{r WinComb1, exercise=TRUE}
allWinDat<- rbind(Winco1,Winco2)

#examine the head and tail to make sure it worked

head(allWinDat,n=3) #first three rows
tail(allWinDat,n=3) #last three rows
```

### Merge with `merge`

Now imagine that we get another dataset which contains the financial data for each location on each day. Let's first look at it. 

```{r Win3, message=FALSE, warning=FALSE}
Winco3<-read.csv("./Messy_CSVs/MergeData/WincoMerge3.csv")[,-1]
```

```{r win3L, exercise=TRUE}
#Examine the Boise data set
head(Winco3)
```

Notice that this data has the same dates and locations as the dataset we just created using `rbind`. We will want to merge the data we just created with this new dataset and make sure that the rows are all in the right place. We can do that with the `merge` command.

```{r winMerge, exercise=TRUE}
WincoFull<-merge(allWinDat,Winco3, by=c("Date","Location"))

#check it out to make sure it worked

head(WincoFull,n=10) #first 10 rows
```

Awesome! We got out datasets all merged!

## Day Two: Dates

## Day Two: Times

## Day Two: Summary and Quiz

### Recap

### Resources

### Follow-up Videos

![](https://www.youtube.com/watch?v=Zc_ufg4uW4U)

### Quiz


## Session Information
```{r sessioninfo, echo=FALSE}

sessionInfo()

```

